---
title: 'Exercises: data validation'
author: "Mark van der Loo"
date: "Statistics Netherlands"
output: 
  pdf_document:
    includes: 
      in_header: ../00tex/header.tex
params:
  answers: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Before we start.

1. Set up a new RStudio project called 'datacleaning'. RStudio will create a new
directory called `datacleaning` (make sure  you can find that directory with your
file browser).
2. Copy the files `SBS2000.csv` and `rules.R` into the `datacleaning` directory.
3. Read the data into your R session using the following code.
```{r}
SBS2000 <- read.csv("SBS2000.csv",stringsAsFactors = FALSE)
```

Use functions such as `head()` and `summary()` to acquaint yourself with this
data. Play around for a few minnutes and make some plots and descriptive
statistics.

#### Note

Always work in an R script, and use `<CTRL>+<ENTER>` to execute your code.
Working in the command-line directly is a bad practice that makes all your work
irreproducible.



#### Exercise 1: Validating data with `validate`

The validate package provides a data validation infrastructure for R. To get
started, try the following code.

```{r, echo=TRUE, results='hide', fig.show='hide'}
library(validate)

# define a ruleset
rules <- validator(
  turnover >= 0
  , other.rev >= 0
  , turnover + other.rev == total.rev
  , if (staff > 0) staff.costs > 0
)
# print the ruleset
rules

# confront data with rules
result <- confront(SBS2000, rules)

# print results
result

# summarise results
summary(result)

# visualize results
plot(result)
```

Also try the following
```{r, echo=TRUE, results='hide', eval=FALSE}
dat <- as.data.frame(result)
View(result)
```
Interpret the results, and answer the following questions.

a. How many records violate the rule `other.rev >= 0`?
b. Take a close look at `summary(result)`. Explain why there are differences 
beteen the `expression` column and the rule that we defined above.
c. Which rule was violated most often?
d. Why are there 37 `NA`s for rule `V3`?


#### Terminology

- **validator.** Rules are stored in an object of class `validator`. 
- **validation.** The results of a `confront`ation are stored in an object
of class `validation`.


#### Exercise 2. Reading rule sets from file

Try the following code (note the `.` in `.file=`).
```{r}
myrules <- validator(.file="rules.R")
```


a. Open the file `rules.R` Think of at least 8 extra rules for this data set,
add them (with comments) to the file; read the rule set and confront the data
with the rules. Make summaries and plots.
b. What rule is violated most often? 
c. Which records violate most rules? See the help files of the
`sort` and `aggregate` methods for validation objects. 
```{r, eval=FALSE}
?`sort.validation-method`
?`aggregate.validation-method`
```
Or if you are familiar with it, use methods from the `dplyr` package to get this
information from `as.data.frame(your result)`.

#### Exercise 3. More rule syntax

In `validate`, any R expression that results in a `logical` (boolean) result
is accepted as a _validation rule_. A concise overview can be found in `?syntax`.


a. Add the following rules to `rules.R`, and analyse the results after confronting them with the data.
    - The average profit must be larger than or equal to zero.
    - The covariance between staff and turnover must be larger than zero (more
    staff is on average associated with more turnover). See `?cov` and choose an
    appropriate manner to handle missing data (parameter `"use="`).
    - `size` must be  one of `"sc0"`, `"sc1"`, `"sc2"`, or `"sc3"` (tip: use the `%in%` operator).
    - `inc.prob` must be nonnegative and not larger than 1.


b. It is possible to check types: any function starting with `is.` is
allowed as validating function, e.g. the following rule checks that
`staff` is numeric.
```{r, eval=FALSE}
is.numeric(staff)
```
Add the folowing rules to `rules.R`, and analyse the results after confronting
them with the data.,
    - Ensure that the numeric variables are actually of type `numeric`.
    - Ensure that `id` is of class `character`.


In `validate`, you can create 'variable groups' for sets of variables that
need to satisfy the same rule. For example, in `rules.R`
```{r, eval=FALSE}
G := var_group(id, size)
is.character(G)
```
is equivalent to
```{r, eval=FALSE}
is.character(id)
is.character(size)
```

c. Use variable groups to check for all numeric data whether they satisfy `is.numeric`.
(i.e. simplyfy the previous code).






In `validate` rules are expressed using variable names as they occur as
column names in the data set. It is also possible to refer to the data set as
a whole, using the "`.`". Syntax. For example, the following rule states that
there must be at least 50 rows.

```{r, eval=FALSE}
nrow(.) >= 50
```

d. Add the following rules to `rules.R` and analyse the results after confronting
them with the data.
    - Number of rows must be larger than or equal to 50
    - Number of columns must equal 11
    - The column `id` must exist (hint: use `names(.)`).
    
    
e. Consider again the Statistical Value Chain, and the rules now present in 
`rules.R`. Which demands apply to `Raw` data, which to `Input`
data and which to `Valid` data? Put the rules into different files accordingly.


#### Exercise 4 (theory). Validation functions

Explain why the following are, or are not data validation functions.

a.
$$
v(x,y) = \left\{\begin{array}{l}
\texttt{TRUE} \textrm{ if } x + y = 3\textrm{ and } x-y = 0\textrm{ and }x \leq1\\
\texttt{FALSE} \textrm{ otherwise}
\end{array}
\right.
$$

b.
$$
v(x) = \left\{\begin{array}{l}
\texttt{TRUE} \textrm{ if } x \geq 3 \textrm{ and } x \geq 5 \\
\texttt{FALSE} \textrm{ otherwise}
\end{array}
\right.
$$

c.
$$
v(x) = \left\{\begin{array}{l}
\texttt{TRUE} \textrm{ if } x \leq 10 \textrm{ or } x \geq 3\\
\texttt{FALSE} \textrm{ otherwise}
\end{array}
\right.
$$

```{r, echo=params$answers}
# a. No. The three demands cannot all be TRUE at the same time, 
#    so v is FALSE everywhere.
# b. Yes, although there is redundancy (we can leave x>=3 out).
# c. No, this ALWAYS results in TRUE so it does not validate anything.
```




#### Exercise 5. Reporting

Use `rmarkdown` to generate a data quality report (pdf of HTML). The report
should be fully automated so it must:

1. Read the data file and read the rules.
2. Confront the data with the rules
3. Show a few tables and plots.

This is a free assignment: use your imagination!





























