---
title: "Systematic Data Cleaning for Official Statistics with R"
author: "Mark van der Loo"
date: "May 8 2019 | INE Spain"
output:
  beamer_presentation:
    fig_caption: false
    latex_engine: xelatex
    keep_tex: false
    includes:
      in_header: ../00tex/presheader2.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Agenda

- R at the office
- Data processing
- R tools for data cleaning
- Detailed examples + conclusions
- Awesome official statistics software

# R at the office

## Introduction of R at statistics netherlands

- `S+` and `R` First Used as research tool at dpt of methodology.
- `R` introduced as strategic tool in 2010
    - Introductory course (methods dpt)
    - User meetings
    - Wiki
    - Code standars, IT standards
- Currently 100+ users, course by stats dpts.


## Use of R at Statistics Netherlands

### Throughout the office. 

- Deriving the business cycle indicator
- Analyzing supply-and-demand tables
- Tourism statistics
- Computing HSMR
- Editing health care institute data
- Animal population statistics
- $\vdots$

\begin{picture}(0,0)
\put(200,20){\includegraphics[height=4cm]{fig/conjunctuurklok.png}}
\end{picture}

## Installation and infrastructure

### Current situation

- Central installation of R and RStudio, accessible to all in P-environment.
- 8-core/32G Remote Desktop Servers for heavier interactive work
- Batch environment for scheduled tasks

### Future situation

- RStudio Server Pro

### In both cases

- Curated set of pre-installed packages
- Extra packages installable via local CRAN repository
- Older versions of R remain accessible after upgrade


## More information


\begin{columns}
  \begin{column}{0.3\textwidth}
    \includegraphics[width=\textwidth]{fig/kowarik2018using.png}
  \end{column}
  \begin{column}{0.5\textwidth}
    A. Kowarik and M. van der Loo. \emph{Romanian 
    Statistical Review} \textbf{1/2018} 15--29.
  \end{column}
\end{columns}






# Data processing

## (part of the) Statistical Value Chain

\begin{center}
\includegraphics[height=3cm]{fig/valuechain.pdf}
\end{center}


### Notes

- This part only pertains to the data processing stage. Collection, design,
dissemination is not included.
- The fixed points are well-defined statistical products.


## (part of the) SVC and GSBPM

\begin{center}
\includegraphics[height=6cm]{fig/GSBPMSVC.png}
\end{center}


## The SVC: Remarks

- Actual data processing is not necessarily linear across the chain
- In production architectures a more flexible model is often used where
the definition of interfaces between processing steps play a crucial role.
The chain shown here is a general example covering most steps in some way.
- The general idea scales really well.

## More information

\begin{columns}
\begin{column}{0.3\textwidth}
\includegraphics[width=\textwidth]{fig/struijs2013redesign.png}
\end{column}
\begin{column}{0.7\textwidth}
P. Struijs, A. Camstra, R. Renssen and B. Braaksma. \emph{Journal of Official Statistics}
\textbf{29/1} 2013 49--71.
\end{column}
\end{columns}



# R tools for data cleaning

## Data cleaning

\begin{center}
\includegraphics[height=2.5cm]{fig/valuechain1.png}
\end{center}


## High-level process view (CSPA, GSIM)


\begin{center}
\includegraphics[height=4cm]{fig/datastep2.pdf}
\end{center}

\begin{center}
Separation of concerns + Modular approach
\end{center}


## Slightly more realistic process view

\begin{center}
\includegraphics[height=0.7\textheight]{fig/dataflow.pdf}
\end{center}

## What does that mean for implementation?

### Our approach

- Separation of concerns $\to$ DSL
- Modularity `==` **UNIX philosophy**
    - each tool doing one thing, really well.
    - tools should integrate with ease
- Design for **humans**
- In **open source**
- **Reuse** existing tools where possible

### R

- Free/Libre and open source
- Well-established, many users
- Very flexible (functional PL)
- Best package system


\begin{picture}(0,0)
\put(220,70){\includegraphics[height=2cm]{fig/R_logo.png}}
\end{picture}

## By the way

\vspace{2cm}

\begin{center}
\emph{R is probably the most thoroughly validated statistics software on Earth.}
\end{center}
\hfill{}\scriptsize{Uwe Ligges, CRAN maintainer (useR!2017)}

\begin{picture}(0,0)
\put(250,60){\includegraphics[height=3.5cm]{fig/uwe.png}}
\end{picture}

## R tools, available on CRAN

\begin{center}
\includegraphics[height=0.7\textheight]{fig/stack.png}
\end{center}


## Implementation: preparation

```{r, eval=FALSE,echo=TRUE}
dat    <- read.csv("data/SBS2000.csv")
rules  <- validate::validator(.file="data/rules.R")
logger <- validate::lbj_rules(rules)
```

\begin{center}
\includegraphics[height=3.5cm]{fig/data_and_rules.png}
\end{center}


## Implementation: data cleaning

```{r,eval=FALSE, echo=TRUE}
dat %L>%                    
    lumberjack::start_log(logger) %L>%  
    errorlocate::replace_errors(rules) %L>%
    tag_missing() %>%
    simputation::impute_mf(. – id ~ . - id) %L>%
    rspa::match_restrictions(rules, eps=1E-8) %L>%
    dump_log() ->    
    clean_data
```

## Results

\begin{picture}(0,0)
\put(0,-20){\includegraphics[height=4cm]{fig/before.png}}
\put(30,-130){\includegraphics[height=4cm]{fig/after.png}}
\put(150,-100){\includegraphics[height=6cm]{fig/rules_log.png}}
\end{picture}


# Detailed examples

## The `validate` package

### Purpose: data checking (validation)

- Define, apply, and maintain data validation rules.
- Summarize, investigate, export validation results.
- Separate domain knowledge from programming skills.

### Example data

```{r, echo=TRUE}
library(validate)
SBS2000 <- read.csv("data/SBS2000.csv")
rules <- validator(
  staff >= 0
  , if (staff > 0) staff.costs > 0
  , turnover + other.rev == total.rev
  , turnover >= 0, other.rev >= 0
)
```

## Confronting data with rules; plotting, summarizing


```{r, echo=TRUE,fig.height=4.5}
result <- confront(SBS2000, rules, key="id")
plot(result)
```


## Confronting data with rules; plotting, summarizing

```{r, echo=TRUE}
summary(result)
```

## Validation results are data

```{r, echo=TRUE}
head( as.data.frame(result) )
```

## Validation rules are also data

\scriptsize{}
```{r, echo=TRUE}
rules
# rules can have rich metadata
rules[[2]]
```
\normalfont{}

## Validation rules can be investigated

```{r, echo=TRUE}
variables(rules)
```

### Are all variables in the data covered by the rules?
```{r, echo=TRUE}
all( variables(SBS2000) %in% variables(rules) )
```

## Package overview

\begin{center}
\includegraphics[height=0.7\textheight]{fig/overview.pdf}
\end{center}

## Example: rules with rich metadata (YAML)

```{}
---
options:
  raise: none
  lin.eq.eps: 1.0e-08
  lin.ineq.eps: 1.0e-08
---
rules:
- expr: turnover >= 0
  name: 'rule-01'
  label: 'nonnegative income'
  description: | 
    'Income cannot be negative (unlike in the
     definition of the tax office)'
  created: 2018-06-05 14:44:06
  origin: rules.txt
  meta: []
```


# Imputation




## Imputation in R

### Specialized packages

  - Many available (VIM, mice, Amelia, mi, $\ldots$)
  - Interfaces vary (a lot)

### DIY with model/predict

```{r, eval=FALSE, echo=TRUE}
m <- lm(Y ~ X, data=mydata)
ina <- is.na(mydata$Y)
mydata[ina, "Y"] <- predict(m, newdata = mydata[ina,])
```

- Code duplication, doesn't always work



\note{
When I set up an imputation system, I think about the problem, implement and 
test it, evaluate and repeat. These widely varying interfaces slow down this
cycle significantly.
}



## Goal of the `simputation` package

### Provide

- a _uniform interface_,
- with _consistent behaviour_,
- across _commonly used methodologies_


### To facilitate

- experimentation 
- configuration for production


## A first example

```{r}
library(magrittr)
library(simputation)
data(SBS2000,package="validate")
```

```{r, echo=TRUE }
dat <- SBS2000[4:7]
dat %>% head(3)
```

```{r, echo=TRUE}
dat %>% impute_lm(other.rev ~ turnover) %>% head(3)
```




## The `simputation` package

### An imputation prodedure is specified by

1. The variable to impute
2. An imputation model
3. Predictor variables



### The simputation interface

```
impute_<model>(data
  , <imputed vars> ~ <predictor vars>
  , [options])
```

## Experimenting with methods

```{r, echo=TRUE}
# linear model
dat %>% impute_lm(other.rev ~ turnover) %>% head(3)
```

```{r, echo=TRUE}
# robost linear model (M-estimator)
dat %>% impute_rlm(other.rev ~ turnover) %>% head(3)
```



## Example (ct'd): chaining methods

```{r, echo=TRUE}
dat %>% 
  impute_rlm( other.rev ~ turnover ) %>%
  impute_rlm( other.rev ~ staff    ) %>%
  head(3)
```


## Example: grouping

```{r, echo=TRUE, eval=FALSE}
SBS2000 %>% impute_rlm(total.rev ~ turnover | size) 

# or, using dplyr::group_by
SBS2000 %>% 
  group_by(size) %>%
  impute_rlm(total.rev ~ turnover)
```


## Example: add random residual

```{r, echo=TRUE, eval=FALSE}
SBS2000 %>% impute_rlm(total.rev ~ turnover | size,
                         add_residual="observed")

SBS2000 %>% impute_rlm(total.rev ~ turnover | size,
                         add_residual="normal") 
```


## Example: proxy imputation

```{r, echo=TRUE, eval=FALSE}
# impute VAT value where turnover is unavailable
SBS2000 %>% impute_proxy(turnover ~ vat)
```

```{r, echo=TRUE, eval=FALSE}
# transformations are also allowed
SBS2000 %>% impute_proxy(other.rev ~ total.rev - turnover)
```

## Example: train on `A`, apply to `B`

```{r,echo=TRUE, eval=FALSE}
m <- MASS::rlm(other.rev ~ turnover + staff
               , data=SBS1999)
impute(SBS2000, other.rev ~ m) 
```




## Currently available methods

- Model based (optional random residual):
    - standard/$M$/elasticnet regression 
    - CART models and Random forest
- Multivariate
    - EM-based imputation
    - missForest (=iterative random forest)
- Donor imputation (including various donor pool specifications)
    - k-nearest neigbour (based on [gower](https://cran.r-project.org/package=gower)'s distance)
    - sequential, random hotdeck
    - Predictive mean matching
- Other
    - (groupwise) median imputation (optional random residual)
    - Proxy imputation: copy another variable or use a simple transformation
      to compute imputed values.


## Tracking changes in data with `lumberjack`

\begin{center}
\includegraphics[]{fig/datastep2.pdf}
\end{center}

\begin{picture}(0,0)
\put(230,-10){\includegraphics[height=2cm]{fig/lumberjack.jpg}}
\end{picture}

## How to compare two versions of a dataset? (1)

Count changes in cells

\begin{center}
\includegraphics[width=0.62\textwidth]{fig/cellsplit.pdf}%
\includegraphics[width=0.38\textwidth]{fig/cellcount.pdf}
\end{center}

## How to compare two versions of a dataset? (2)

Count changes in rule violations

\begin{center}
\includegraphics[width=0.62\textwidth]{fig/rulesplit.pdf}%
\includegraphics[width=0.38\textwidth]{fig/rules_log.png}
\end{center}


## How to compare two versions of a dataset? (3)

### There are more ways

- Comparing cell-by cell
- Following summary statistics (e.g. means)
- $\vdots$

## The `lumberjack` package

- Works for _any_ data transformation function (no special protocols required)
- What to log is _user-definable_, and _extensible_
- Near-zero change in common R workflow

## Simple example

```{r}
library(lumberjack)
```

```{r,echo=TRUE}
SBS2000 %L>%
  start_log(cellwise$new(key="id")) %L>%
  impute_const(other.rev ~ 0) %L>%
  dump_log() -> imputed_data
```

\scriptsize{}
```{r,echo=TRUE}
read.csv("cellwise.csv") %>% head(3)
```
\normalfont{}

# Conclusions 

## Lessons learned

### (1) Modularity 

Requires careful research. Basic actions (like data validation, logging) preferably defined with 
mathemtical precision.


### (2) Separation of concerns, building for humans

- Follows, if (1) has been thought through well enough.
- Building something that _lasts_ still requires careful thinking, and sometimes new ideas.

### (3) Implementing

- Using FOSS tools (here: R) makes the design $\to$ build $\to$ test quick and easy.
- The fact that R is a functional programming language offers abstractions that are much harder
to build in non-functional languages.
- Still needs deep thinking sometimes.

## The Dolly Parton principle

\begin{center}
\includegraphics[height=6cm]{fig/dolly.jpg}
\end{center}




# The awesomelist of official statistics software

## What are awesomelists?

\begin{center}
\includegraphics[height=4cm]{fig/awesome.png}
\end{center}

### Awesomlist

- A _curated_ list of things, considered _AWESOME_ by the curator(s)
- Hosted on github

\hfill{}\scriptsize{Sindre Sorhus, github.com/sindresorhus/awesome}

## Awesome official statistics software

### Official statistics software is _AWESOME_ if and only if

1. It is free, open source and available for download
2. It is confirmed to be used in the production of official statistics by at least one institute _or_
3. it provides access to official statistics publications

### Also

_We prefer packages that are reasonably easy to install and use, that have at least one stable version, and that are actively maintained._

\vspace{1ex}

\hfill{}\scriptsize{}{O. ten Bosch and M. van der Loo, www.awesomeofficialstatistics.org}


## Awesome official statistics software


\begin{center}
\includegraphics[height=6cm]{fig/awesomelist.png}
\end{center}

## Parts of GSBPM covered

\begin{center}
\includegraphics[height=6cm]{fig/awesomeGSBPM.png}
\end{center}


## If you are awesome, you get to wear the _BADGE_

\begin{center}
\includegraphics[height=6cm]{fig/vim.png}
\end{center}

\hfill{}\scriptsize{M. Temple \emph{et al.} https://github.com/statistikat/VIM\phantom{fooobaar}}


## More information


\begin{columns}
  \begin{column}{0.5\textwidth}
    \includegraphics[height=0.7\textheight]{fig/sdcr.jpg}%
  \end{column}
  \begin{column}{0.5\textwidth}
    \textbf{Contact}
    \begin{itemize}
    \item @markvdloo
    \item github.com/markvanderloo
    \item mplo@cbs.nl
    \end{itemize}
    $$
    \;
    $$
  
  
   \textbf{Book}
    MPJ van der Loo and E de Jonge (2018) John Wiley \& Sons.
   $$
   \;
   $$
   \textbf{More FOSS}
   \href{www.awesomeofficialstatistics.org}{www.awesomeofficialstatistics.org}
   
    
  \end{column}
\end{columns}








