---
title: "Easy imputation with the simputation package"
author: "Mark van der Loo"
date: ""
output: 
  beamer_presentation:
    fig_caption: false
    includes:
      in_header: tex/header.tex
header-includes:
- \titlegraphic{\vspace{-1.5cm}\includegraphics[width=4cm]{fig/useR2017logo.jpg}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(simputation)
library(magrittr)
library(lumberjack)
```


## Example data

```{r}
data(retailers, package='validate')
ret <- retailers[3:7]
head(ret, 3)
```


\begin{picture}(0,0)
\put(266,123){\includegraphics[height=1.5cm]{fig/shopping-cart.png}}
\end{picture}



## Imputation in R

### Specialized packages

  - Many available (VIM, mice, Amelia, mi, $\ldots$)
  - Interfaces vary (a lot)

### DIY with model/predict

```{r, eval=FALSE, echo=TRUE}
m <- lm(Y ~ X, data=mydata)
ina <- is.na(mydata$Y)
mydata[ina, "Y"] <- predict(m, newdata = mydata[ina,])
```

- Code duplication, doesn't always work

\begin{picture}(0,0)
\put(260,150){\includegraphics[height=1.8cm]{fig/work.png}}
\end{picture}


\note{
When I set up an imputation system, I think about the problem, implement and 
test it, evaluate and repeat. These widely varying interfaces slow down this
cycle significantly.
}

## Idea of the simputation package

### Provide

- a _uniform interface_,
- with _consistent behaviour_,
- across _commonly used methodologies_


### To facilitate

- experimentation 
- configuration for production

\begin{picture}(0,0)
\put(270,120){\includegraphics[height=1.8cm]{fig/lightbulb.png}}
\end{picture}



## The simputation interface

```
impute_<model>(data
  , <imputed vars> ~ <predictor vars>
  , [options])
```

### Example: linear model imputation

```{r}
impute_lm(ret, other.rev ~ turnover) %>% head(3)
```

\note{
For any imputation there are three things to specify:
\begin{itemize}
\item The method used
\item The imputed variable
\item The predictor variables
\end{itemize}
}

\begin{picture}(0,0)
\put(270,170){\includegraphics[height=1.5cm]{fig/pushbutton.png}}
\end{picture}

## Example: chaining imputations

```{r}
ret %>% 
  impute_lm(other.rev ~ turnover + staff) %>% 
  impute_lm(other.rev ~ staff) %>%
  head(3)
```

\begin{picture}(0,0)
\put(270,130){\includegraphics[height=1.8cm]{fig/gears.png}}
\end{picture}



## Example: robust imputation ($M$-estimation) 

```{r}
ret %>% 
  impute_rlm(other.rev ~ turnover + staff) %>% 
  impute_rlm(other.rev ~ staff) %>%
  head(3)
```

\begin{picture}(0,0)
\put(270,130){\includegraphics[height=1.8cm]{fig/gears.png}}
\end{picture}


## Example: Multiple variables, same predictors

```{r,eval=FALSE}
ret %>% 
  impute_rlm(other.rev + total.rev ~ turnover) 

ret %>% 
  impute_rlm( . - turnover ~ turnover) 
```

\begin{picture}(0,0)
\put(270,80){\includegraphics[height=1.5cm]{fig/copy.png}}
\end{picture}


## Example: grouping

```{r, eval=FALSE}
retailers %>% impute_rlm(total.rev ~ turnover | size) 

# or, using dplyr::group_by
retailers %>% 
  group_by(size) %>%
  impute_rlm(total.rev ~ turnover)
```

\begin{picture}(0,0)
\put(270,130){\includegraphics[height=1.4cm]{fig/groups.png}}
\end{picture}

## Example: add random residual

```{r, eval=FALSE}
retailers %>% impute_rlm(total.rev ~ turnover | size,
                         add_residual="observed")

retailers %>% impute_rlm(total.rev ~ turnover | size,
                         add_residual="normal") 
```

\begin{picture}(0,0)
\put(260,120){\Huge$\boldsymbol{+\varepsilon}$}
\end{picture}


## Example: train on `A`, apply to `B`

```{r,eval=FALSE}
m <- MASS::rlm(other.rev ~ turnover + staff
               , data=retailers)
impute(ret, other.rev ~ m) 
```

\begin{picture}(0,0)
\put(270,80){\includegraphics[height=1.5cm]{fig/recycle.png}}
\end{picture}

## Currently available methods


- Model based (optional random residual):
    - standard/$M$/elasticnet regression 
    - CART models and Random forest
- Multivariate
    - EM-based imputation
    - missForest (=iterative random forest)
- Donor imputation (including various donor pool specifications)
    - k-nearest neigbour (based on [gower](https://cran.r-project.org/package=gower)'s distance)
    - sequential, random hotdeck
    - Predictive mean matching
- Other
    - (groupwise) median imputation (optional random residual)
    - Proxy imputation: copy another variable or use a simple transformation
      to compute imputed values.

\begin{picture}(0,0)
\put(260,160){\includegraphics[height=2.0cm]{fig/toolkit.png}}
\end{picture}


## Who imputed what? Ask the lumberjack!

### Lumberjack

A _pipe operator_ that _logs changes in data_.

### Provides

- `%>>%`: the lumberjack operator
- `start_log()`: start loggin'
- `dump_log()` : dump log to file
- `stop_log()` : end loggin'
- Several loggers

### Fully extendable
- Users can provide their own loggers

\begin{picture}(0,0)
\put(275,150){\includegraphics[height=2.3cm]{fig/lumberjack.jpg}}
\end{picture}



## Example

\scriptsize
```{r}
ret$id <- seq_len(nrow(ret))
logger <- cellwise$new(key="id")
imputed <- ret %>>%
  start_log(logger) %>>%
  impute_rlm(other.rev ~ total.rev + staff) %>>%
  impute_median(other.rev ~ 1) %>>%
  dump_log(stop=TRUE)
read.csv("cellwise.csv") %>>% head(3)
```
\normalsize



## Some pointers

### Getting started
```{r,eval=FALSE}
install.packages('simputation', dependencies = TRUE)
install.packages('lumberjack')
```
```{r, eval=FALSE}
vignette("intro",package="simputation")
vignette("intro",package="lumberjack")
```

### Code / issues:
 \href{https://github.com/markvanderloo/}{\texttt{github.com/markvanderloo/}}

### Contact
 \texttt{@markvdloo} | \texttt{mark.vanderloo@gmail.com}

\begin{picture}(0,0)
\put(291,166){
\includegraphics[height=0.9cm]{fig/download.png}
}
\put(291,119){
\includegraphics[height=0.9cm]{fig/on.png}
}
\put(291,62){
\includegraphics[height=1cm]{fig/bug.png}
}
\put(290,18){
\includegraphics[height=0.8cm]{fig/contact.png}
}
\end{picture}




