---
title: "From Official Statistics to Official Data Science"
author: "Mark van der Loo, Statistics Netherlands"
date: "Complutense University of Madrid, Spring 2019"
output:
  beamer_presentation:
    fig_caption: false
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---




## Agenda


1. Why are computing skills important? 
    - Some personal observations.
    - Experiences as a research methodologist
2. Official Statistics as a (Data) Science

# Observations

## Example one

Methodologist specifies

$$
\textrm{mean}(x) = \frac{x_1}{\pi_1} +\frac{x_2}{\pi_2}+\cdots+\frac{x_n}{\pi_n}
$$

Software developer implements

```{r,eval=FALSE}
sum(x)/3.14
```

## Example two


Methodologist specifies

$$
\textrm{geometric\_mean}(x) = \sqrt[n]{x_1 \times x_2\times\cdots \times x_n}
$$

Software developer implements

```{r}
geom_mean = function(x) prod(x)^(1/length(x))
```

## Example two (continued)

Software developer tests implementation
```{r}
geom_mean(c(4,4)) == sqrt(16)
```

User puts some actual data in: $1,2,\ldots,200$
```{r}
geom_mean(1:200)
```

##

\begin{center}
\includegraphics[height=6.5cm]{fig/onedoes.jpg}
\end{center}


## Lessons learned

### Implementing methods is _not trivial_

It is called _scientific computing_ or _numerical mathematics_, and it is a scientific field.



### For (project) management in particular

You need to be able to recognize these situations to put the right person on the job.


\begin{picture}(0,0)
\put(200,-60){
\includegraphics[height=2cm]{fig/nummath.jpg}%
\includegraphics[height=2cm]{fig/numrec.jpg}%
}
\end{picture}


## A question to statistics managers

Your 'computer person' retires or leaves. You need to hire someone that will
modernize the systems developed by this person.

a. What do you put in the job advertisement?
b. How do you interview this person to asses maturity in (statistical) programming?

## A question for strategic management

### Core question

Do you think that statistical computing is a core competence for the statistical office?

### and if so,

How much of it is needed (FTE)? Should there be associated career paths? $\ldots$


# Experiences as a research methodologist


## High-level process view (CSPA, GSIM)


\begin{center}
\includegraphics[height=4cm]{fig/datastep2.pdf}
\end{center}

\begin{center}
Separation of concerns + Modular approach
\end{center}


## Slightly more realistic process view

\begin{center}
\includegraphics[height=0.7\textheight]{fig/dataflow.pdf}
\end{center}


## Data cleaning using R-based packages (1)

```{r, eval=TRUE, echo=TRUE}
library(validate)
SBS2000 <- read.csv("SBS2000.csv")
rules   <- validator(.file = "rules.R")
```

\begin{center}
\includegraphics[height=3.5cm]{fig/data_and_rules.png}
\end{center}

## Data cleaning using R-based packages (2)

```{r, fig.height=4}
out <- confront(SBS2000, rules)
plot(out)
```


## Data cleaning using R-based packages (3)

```{r,echo=FALSE}
set.seed(2)
```

```{r, message=FALSE, results='hide'}
library(lumberjack); library(rspa); 
library(simputation); library(errorlocate)

SBS2000 %L>%                    
    start_log( cellwise$new(key="id") ) %L>%  
    replace_errors( rules ) %L>%
    tag_missing() %L>%
    impute_mf( . - id ~ . - id ) %L>%
    match_restrictions( rules, eps=1E-8 ) %L>%
    dump_log() -> clean_data
```

\begin{picture}(0,0)
\put(250,50){}
\end{picture}

## Data cleaning using R-based packages (3)


```{r, eval=FALSE, results='hide'}
library(lumberjack); library(rspa); 
library(simputation); library(errorlocate)

SBS2000 %L>%                    
    start_log( cellwise$new(key="id") ) %L>%  
    replace_errors( rules ) %L>%
    tag_missing() %L>%
    impute_mf( . - id ~ . - id ) %L>%
    match_restrictions( rules, eps=1E-8 ) %L>%
    dump_log() -> clean_data
```

\begin{picture}(0,0)
\put(250,50){\includegraphics[height=2.5cm]{fig/datastep2.pdf}}
\end{picture}


## Data cleaning using R-based packages (4)

```{r, fig.height=4}
out <- confront(clean_data, rules)
plot(out)
```


## Data cleaning using R-based packages (5)

\scriptsize{}
```{r}
read.csv("cellwise.csv") %L>% head(3)
```
\normalfont{}

## What went into this?

### Methodology

Calculus, linear algebra, algorithm design, (convex) optimization, linear programming, formal logic,
mathematical modeling.

### Implementation

Parsing and language theory, functional programming, object orientation,
numerical methods,  algebraic data types. LOTS of programming
experience, compiled languages, APIs and technical standards. Also: version
control, documenting and testing, CI tools, UX design.







## The Dolly Parton Principle


### Dolly

It takes a lot of money to look so cheap.



### Me, writing software

It takes a lot of thinking to look so simple.

\begin{picture}(0,0)
\put(220,-20){\includegraphics[height=4cm]{fig/dolly.jpg}}
\end{picture}


# Official Statistics as a (Data) Science

## Data science skill set

\begin{center}
\includegraphics[height=6.5cm]{fig/dsvenn.jpg}
\end{center}
\hfill{}\scriptsize{Nolan and Temple Lang (2010) \emph{The American Statistician} \textbf{64}(2) 97--107}


## Data science skill set

\begin{center}
\includegraphics[height=6.5cm]{fig/dsvenn2.png}
\end{center}
\hfill{}\scriptsize{Drew Conway (2013) \underline{\href{http://drewconway.com/zia/2013/3/26/the-data-science-venn-diagram}{blog post}}}

## Data science skill set?

\begin{center}
\includegraphics[height=6.5cm]{fig/dsvenn3.pdf}
\end{center}
\hfill{}\scriptsize{Me, reproduced from memory as seen at \emph{The Internets} }

## Types of data scientists

\begin{center}
\includegraphics[height=6.5cm]{fig/dsradar.png}
\end{center}
\hfill{}\scriptsize{Mango Solutions Data Science Radar}


## Data Science

\begin{quote}
Science of planning for, acquisition, management, analysis of, and inference
from data.
\end{quote}

\hfill{}\scriptsize{StatNSF (2014); De Veaux \emph{et al} 2017 \emph{Annu.~Rev.~Stat.} \textbf{4} 15--31}

## Is data science a science?


\begin{quote}
[...] there is a solid case for some entity called ‘Data Science’ to be
created, which would be a true science: facing essential questions of a lasting
nature and using scientifically rigorous techniques to attack those questions
\end{quote}

\hfill{}\scriptsize{Donoho (2015) \emph{50 years of data science}.}


## Key competencies of a data science major

1. Computational and statistical thinking
2. Mathematical foundations
3. Model building and assessment
4. Algorithms and software foundation
5. Data curation
6. Knowledge transference---communication and responsibility

\hfill{}\scriptsize{De Veaux \emph{et al} 2017 \emph{Annu.~Rev.~Stat.} \textbf{4} 15--31}


## Curriculum

\begin{center}
\includegraphics[height=6.5cm]{fig/veauxcurr.png}
\end{center}
\hfill{}\scriptsize{De Veaux \emph{et al} 2017 \emph{Annu.~Rev.~Stat.} \textbf{4} 15--31}



## Extra subject areas of an official statistics major

1. Macroeconomics
2. Demography
3. Ontologies and metadata
3. Policy, governance, international context
4. Privacy and data safety


## Mark's Official Data Science Bachelors Curriculum


\begin{center}
{\includegraphics[height=7cm]{fig/ects.pdf}}
\end{center}


## Semester I

- Calculus (6 ECTS)
    - \scriptsize{Set theory, calculus on the real line, investigating functions (min, max, asymptotes), multivariate calculus, Lagrange multiplier method}
- Linear algebra (6 ECTS)
    - \scriptsize{Vectors and vector spaces, linear systems of equations and matrices, matrix inverse, eigenvalues, inner product spaces.}
- Introduction to programming (4 ECTS)
    - \scriptsize{Imperative programming, algorithm design, recursion, complexity, practical assignments.}
- Public policy and administration (4 ECTS)
    - \scriptsize{Government structure and institutions, policy-making and implementation, role of official statistics, international context, privacy}

\begin{picture}(0,0)
\put(200,-40){
\includegraphics[height=2cm]{fig/calculus.jpg}%
\includegraphics[height=2cm]{fig/linalg.jpg}%
\includegraphics[height=2cm]{fig/algorithms.jpg}
}
\end{picture}

    
## Semester II

- Probability and statistics I (6 ECTS)
    - \scriptsize{Probability, discrete and continuous distributions, measures of location and variation, Bayes' rule, sampling distributions, estimation of mean and variance, CLT, ANOVA, linear models}.
- Linear programming and optimization (4 ECTS)
    - \scriptsize{Recognizing and modeling LP problems, simplex method, duality, sensitivity analysis, intro nonlinear optimization. Practical assignments using software tools.}
- Programming with data I (4 ECTS)
    - \scriptsize{Statistical analysis, data visualisation and reporting, programming skills and reproducibility, version control, testing, project.}
- Macroeconomics (6 ECTS)
    - \scriptsize{National Accounts, economic growth, labour market, consumption and investments, inflation, macro-economic equilibrium, budget policy and government debt. The main surveys.}

\begin{picture}(0,0)
\put(200,-40){
\includegraphics[height=2cm]{fig/macroecon.jpg}%
\includegraphics[height=2cm]{fig/probstat.jpg}%
\includegraphics[height=2cm]{fig/linprog.jpg}%
\includegraphics[height=2cm]{fig/r4ds.png}
}
\end{picture}

## Semester III

- Models in computational statistics (6 ECTS)
    - \scriptsize{GLM, regularization, Tree models, Random Forest, SVM, unsupervised learning, model selection, lab with practical assignments.}
- Probability and statistics II (4 ECTS)
    - \scriptsize{Bayesian inference, Gibbs sampling and MCMC, maximum likelihood and Fisher information, latent models}
- Programming with data II (4 ECTS)
    - \scriptsize{Relational algebra and data bases, data representation, regular expressions, and technical standards, ontologies and metadata, practical assignments.}
- Demography (6 ECTS)
    - \scriptsize{Fertility, mortality, life table and decrement processes, age-specific rates and probabilities,
    stable and nonstable population models, cohorts, data and data quality. The main surveys.}

\begin{picture}(0,0)
\put(200,-40){
\includegraphics[height=2cm]{fig/statlearn.jpg}%
\includegraphics[height=2cm]{fig/probstat.jpg}%
\includegraphics[height=2cm]{fig/sdcr.jpg}%
\includegraphics[height=2cm]{fig/demography.jpg}%
}
\end{picture}

## Semester IV

- Methods for official statistics I (4 ECTS)
    - \scriptsize{Advanced survey methods, weighting and estimation, calibration, SAE, handling non-response}
- Methods for official statistics II (4 ECTS)
    - \scriptsize{Time series, seasonal adjustment, benchmarking and reconciliation, time series models}
- Programming with data III (4 ECTS)
    - \scriptsize{Infrastructure for computing with big data, map-reduce, key-value stores, project.}
- Communication (4 ECTS)
    - \scriptsize{Scientific and technical writing, principles of visualization, dissemination systems.}
- Ethics and philosophy of science (2 ECTS)

\begin{picture}(0,0)
\put(120,-40){
\includegraphics[height=2cm]{fig/sarndal.jpg}%
\includegraphics[height=2cm]{fig/nonresponse.jpg}%
\includegraphics[height=2cm]{fig/timeseries.jpg}%
\includegraphics[height=2cm]{fig/benchmarking.jpg}%
\includegraphics[height=2cm]{fig/bigdata.jpg}%
\includegraphics[height=2cm]{fig/philo.jpg}%
}
\end{picture}

## Semester V

- Methods for official statistics III (4 ECTS)
    - \scriptsize{Principles of data editing, Fellegi-Holt error localization, methods for imputation.}
- Methods for official statistics IV (4 ECTS)
    - \scriptsize{Information Security and Statistical Disclosure Control}
- Research methods in social science (4 ECTS)
    - \scriptsize{Questionnaire design and field research, measurement models and latent variables}
- Elective course (4 ECTS)
    - \scriptsize{In the area of social science, economics, econometrics, computer science, or math\&statistics}
- Large programming project (4 ECTS)
    - \scriptsize{E.g. a small production system, a dashboard, data cleaning system}

\begin{picture}(0,0)
\put(180,-40){
\includegraphics[height=2cm]{fig/editing.jpg}%
\includegraphics[height=2cm]{fig/sdcr.jpg}%
\includegraphics[height=2cm]{fig/sdc.jpg}%
}
\end{picture}


## Semester VI

- Elective courses (8 ECTS)
    - \scriptsize{Preparing for thesis research}
- Bachelor's thesis (12 ECTS)
    - \scriptsize{Research in Macroeconomy, Demography, or Methodology. Preferably at an
    NSI or international organization. }


## Some interesting research areas

### Methodology

- Complexity theory, econophysics, agent-based modeling
- Network theory
- Streaming data

### Content / output

- Beyond GDP
- Globalization, regionalization
- SDG, energy transition



## Take-home message

\begin{center}
\Large Official statistics is a (data) \emph{science}, applied to society.
\end{center}



