---
title: 'Exercises: deterministic data correction'
author: "Mark van der Loo"
date: "Statistics Netherlands"
output: 
  pdf_document:
    includes: 
      in_header: ../00tex/header.tex
params:
  answers: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Exercise 1. The cellwise logger

Try the following code.
```{r, results='hide', warning=FALSE}
library(lumberjack)
library(dplyr)
SBS2000 <- read.csv("SBS2000.csv")
logger <- cellwise$new(key="id")

out <- SBS2000 %L>%
  start_log(logger) %L>%
  mutate(ratio=profit/total.rev) 

View(logger$logdata())
```

*Exercise:* Use `%L>%` and `filter` to remove records with `ratio > 10E3` from `out`. Look at the logging data again. 

We now remove the logger.
```{r}
rm(logger)
```


#### Excercise 2. The `cells` logger

Other packages, like `validate`, can define loggers for `lumberjack`. The `cells` logger summarizes
changes in cells.
```{r}
library(validate)
logger <- lbj_cells()
```

Look at the following data processing pipeline.
```{r}
library(errorlocate)
library(simputation)
rules <- validator(.file="ruleset.R")
SBS2000 %>%
  replace_errors(rules) %>%
  impute_rlm(turnover ~ profit) %>%
  impute_rlm(turnover ~ staff) -> output
```

**Exercise:** Adapt the above code so all changes and imputations are logged. Make sure you use `dump_log()` to
write the logging data to `csv` and read the data in afterwards.

### Exercise 3 (open exercise). 

Clean up `turnover`, `total.costs` and `profit` from the SBS2000 dataset. A few pointers:

- Define a few meaningful rules
- You can use everything you have learned before. 

Compute the average turnover, profit and total costs after cleanup.

Try different imputation methods, error localization weights, and so on. What is their influence on the outcome?



