---
title: "Monitoring data cleaning processes"
author: "Mark van der Loo, Statistics Netherlands"
date: "Complutense University of Madrid, Spring 2019"
output:
  beamer_presentation:
    fig_caption: false
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---

## Process overview

\begin{center}
\includegraphics[]{fig/datastep.pdf}
\end{center}


## Process overview

\begin{center}
\includegraphics[]{fig/datastep2.pdf}
\end{center}

## How to measure changes? 1. Track cells

\begin{center}
\includegraphics[height=60mm]{fig/cellsplit.pdf}
\end{center}

```{r,eval=FALSE}
validate::cells()
```

## How to measure changes? 2. Track validations

\begin{center}
\includegraphics[height=60mm]{fig/rulesplit.pdf}
\end{center}

```{r,eval=FALSE}
validate::compare()
```

## How to measure changes between `data` and `data'`?

### Many ways

- List every change (record, variable, old, new) ('diff')
- Count differences in cells, or validation
- $\vdots$
- Note if something has changed (`TRUE`/`FALSE`)

## Needs

### Logging framework

- Supporting any type of comparison of `data` and `data'`
- Supporting any kind of transformation between `data` and `data'`
- Without demanding changes in the transforming functions
- That does not get in the way of the user

## Logging framework

### Idea

- A data cleaning procedure is implemented as a sequence of expressions (a script).
- These expressions are _composed_ into a programe when you run the script (`source()`)
- To obtain a logging framework that is not intrusive for the user, we can _change the way expressions are composed_.

## The `lumberjack` package: preparation

\scriptsize{}
```{r}
dat <- read.csv("SBS2000.csv", stringsAsFactors = FALSE)
head(dat,3)

library(validate)
rules <- validator(.file="ruleset.R")

library(lumberjack)
logger <- cellwise$new(key="id")
```
\normalfont{}

## The `lumberjack` package: clean up

```{r}
dat %L>%
  lumberjack::start_log(logger) %L>%
  errorlocate::replace_errors(rules) %L>%
  rspa::tag_missing() %L>%
  simputation::impute_rhd(. ~ 1, backend="VIM") %L>%
  rspa::match_restrictions(rules) %L>%
  lumberjack::dump_log() -> dat_out
```

## Read the log:


\scriptsize{}
```{r}
read.csv("cellwise.csv") %L>% head(3)
```
\normalsize{}

## Background

The pipe is a sort of _function composition_ operator.
```{r,eval=FALSE}
# Pseudocode:
`%>%` <- function(x, fun){
  return( fun(x) )
}
```

The `lumberjack` does some extra things:
```{r, eval=FALSE}
# Pseudocode
%L>% <- function(x, fun){
  y <- fun(x)
  if ( logger_attached_to(x) ){
    logger <- get_logger(x)
    logger$add_difference(x,y)
  }
  return(y)
}
 
```








