---
title: 'Exercises: deterministic data correction'
author: "Mark van der Loo"
date: "Statistics Netherlands"
output: 
  pdf_document:
    includes: 
      in_header: ../00tex/header.tex
params:
  answers: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Exercise 1. Data correction with `dcmodify` and comparing data versions

The `dcmodify` package lets you define and administer data modifying rules,
much in the same way you can define and use data validation ruls in the `validate`
package.

Try the following code.

```{r}
library(dcmodify)
SBS2000 <- read.csv("SBS2000.csv",stringsAsFactors = FALSE)
out <- modify_so(SBS2000
      , if ( other.rev < 0 ) other.rev <- -1*other.rev)
```

Or, if you like the 'pipe' interface of the `magrittr` package, you can
do 
```{r}
library(dcmodify)
library(magrittr)
out <- SBS2000 %>%
  modify_so( if (other.rev < 0) other.rev <- -1*other.rev )
```

We can use the function `cells` of the validate package to compare the two 
versions of our data.
```{r, results='hide'}
library(validate)
cells(SBS2000, out)
```

a. Read the helpfile of `?cells` and interpret the output.


You can create a `modifier` by reading a rule set from file, just like you can
do it in `validate`:

```{r, eval=FALSE}
mod <- modifier(.file="/path/to/your/file.R")
```

All rules in `dcmodify` are of the form
```{r, eval=FALSE}
if (condition) do something
```
It is possible to have nested if-statements:
```{r, eval=FALSE}
if (condition1){
  if (condition2){
    do something
  }
}
```
Currently, `if-else` statements are not yet supported.

b. Create a file called `modifications.R`, and state in it the following rules.
    - If `other.rev` is negative, flip the sign (as in the example)
    - If `other.rev` is missing, make it equal to zero.
    - If `turnover + other.rev` does not equal `total.rev`, and the difference
    is positive: assign the difference to `other.rev`
c. Apply the rules in `modifications.R` to `SBS2000`. Compare the difference between
in- and output with `cells`.
d. Do you think that the order of rule execution is important for the result? Argue
why.

#### Exercise 2. comparing data versions using rule sets

The function `compare` of the `validate` package allows you to compare
versions of a dataset with respect to a rule set. Try the following code.

```{r, results='hide'}
rules <- validator(turnover >= 0, other.rev>=0, turnover + other.rev>=total.rev)

out <- modify_so(SBS2000, if (other.rev < 0) other.rev <- -1*other.rev)
compare(rules, SBS2000, out)
```

**Exercise:** Read the help file of `?compare` to interpret the result. Which 
validations have changed? What rules and records are involved?


#### Exercise 3: combining steps.

We would like the SBS2000 dataset to satisfy the following rules.

```{r,eval=FALSE}
turnover >= 0
other.rev >= 0
turnover + other.rev >= 0
total.costs >= 0
turnover - total.costs == profit
```

In this exercise we will repair a number of flaws using deterministic methods.
In particular, you can use

- the `validate` package
- `modify`, or `modify_so` from the `dcmodify` package
- `correct_typos`, `impute_lr` from the `deductive` package.
- `replace_errors` from the `errorlocate` package.

**Exercise** Use the above functions and try to improve the data set step by
step (you will probably not be able to fix everything, that's ok). You can use
`validate::compare` or `validate::cells` to monitor changes as you create
versions of your dataset. While you experiment, try to think the following
questions:

  - What is a good order for the cleaning steps?
  - Should the number of passing validations strictly increase at every step? Is it 
    'allowed' to decrease the number of passing validations in a step? If so, under what circumstances?












