---
title: "Methods for deterministic correction"
author: "Mark van der Loo, Statistics Netherlands"
date: "Complutense University of Madrid, Spring 2019"
output: 
  beamer_presentation:
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---

## General strategy in data correction

- Use information stored as validation rules, and information from
    0. systematic domain knowledge (correction rules)
    1. faulty cells (e.g. spelling errors)
    2. other, correct cells (deriving values)
    3. other records (imputation)
    4. other files/sources (manual correction)

## Correction based on domain knowledge

### Example

- Irrelevant fields are often left open (in stead of filing zero)
- Costs are sometimes reported negative (but can be made positive)
- A surplus on a balance can be moved to a 'rest' post.

### Common idea

`IF` some data condition holds `THEN` perform a standard activity.


## Domain knowledge should be separated from process flow

\begin{center}
\includegraphics[]{fig/datastep.pdf}
\end{center}


## `dcmodify`: externalize domain knowledge

```{r, results='hide'}
library(dcmodify)
SBS2000 <- read.csv("SBS2000.csv", stringsAsFactors = FALSE)
mod <- modifier(
    if (other.rev < 0) other.rev <- -1 * other.rev
  , if (staff.costs/staff > 1000) staff.costs <- staff.costs/1000
)
modify(SBS2000, mod)
```


## `dcmodify`: externalize domain knowledge


### Similar to `validate` 

- Read rules from CLI or file
- Rules with metadata
- Apply rules


\begin{picture}(0,0)
\put(180,20){\includegraphics[]{fig/dcmodify.pdf}}
\end{picture}


# Deductive correction

## Deductive correction


### Deductive correction

- A set of methods specific to error circumstances
- Given the ruleset and the faulty data, can we reconstruct what went wrong?

## Method: Spelling errors in numbers.

### Observed data:

\begin{tabular}{rr}
turnover &  1024 \\
costs    &  435  \\
\hline
profit   &  598
\end{tabular}

### Note

- $1024-435 = 589\not=598$ (so error) 
- The difference $598 - 589$ = 9 is divisable by 9.

### Proposition

Given a balance rule $\sum_ix_i - t=0$. If for some record $(\boldsymbol{x},t)$
the value $\sum_{i}x_i-t$ is divisable by 9, then the record can be repaired
by transposition of two digits in one of the variables $x_i$ or $t$.

\hfill{}\scriptsize{} Scholtus (2009) Discussion paper 09049 (CBS)\normalsize{}

## Algorithm (sketch)

- Input: integer record, set of linear balance restrictions
- For each variable in each violated rule
    - Solve for that variable
    - Check if the solution is a typing error away from the original

\hfill{}\scriptsize{}SDCR \S9.3\normalsize{}

### A single typing error

deletion, insertion, or substitution of a digit, or transposition
of two adjacent digits.


## Application

```{r}
library(validate)
library(deductive)

d <- data.frame(turnover = 1024, costs = 435, profit = 598)
rules <- validator(turnover - costs == profit)
correct_typos(d, rules)

```

\begin{picture}(0,0)
\put(180,-20){\includegraphics[]{fig/correct_typos.pdf}}
\end{picture}



## Method: deductive imputation


### Observed data:

\begin{tabular}{rr}
turnover &  1024 \\
costs    &  NA  \\
\hline
profit   &  589
\end{tabular}

### Note

We can derive the value of _costs_ using the rule 

```
turnover - costs == profit
```

## Method: deductive imputation

### Observed costs:

\begin{tabular}{rr}
housing   &  1024 \\
transport &  NA   \\
staff     &  NA   \\
interest  &  300  \\
\hline
total     &  1324 
\end{tabular}

### Observe

Since all variables must be nonnegative, the only possible value for
_transport_ and _staff_ is zero.


# Deductive imputation

## Method: deductive imputation

```{r}
library(validate)
library(deductive)

d <- data.frame(housing=1024, transport=NA, staff=NA, interest=300
              , total=1324)
rules <- validator(housing >= 0, transport >= 0, staff >=0, interest >=0
                   , housing + transport + staff + interest == total )
impute_lr(d,rules)
```


## General method for deductive imputation (sketch)

Set of restrictions $\boldsymbol{Ax}=\boldsymbol{b}$. Split according
to observed and missing:

$$
[\boldsymbol{A}_o,\boldsymbol{A}_m]\left(\begin{array}{c}
\boldsymbol{x}_o\\ \boldsymbol{x}_m
\end{array}\right)
=\boldsymbol{b}
$$
Then, depending on the number and structure of missings, some elements of $\hat{\boldsymbol{x}}_m$
are uniquely determined.
$$
\hat{\boldsymbol{x}}_m = \boldsymbol{A}^+(\boldsymbol{b}-\boldsymbol{A}_o\boldsymbol{x}_o)
+(\boldsymbol{1}-\boldsymbol{A}^+_m\boldsymbol{A}_m)\boldsymbol{w}
$$

- $\boldsymbol{A}^+$: Moore-Penrose inverse of $\boldsymbol{A}$
- $\boldsymbol{w}\in\mathbb{R}^{|m|}$ ($|m|$ is nr of missings in $\boldsymbol{x}$).

\hfill{}\scriptsize{}SDCR \S9.3\normalfont{}

## The good news

### R package `deductive`

`deductive::impute_lr` tries all these methods iteratively untill nothing 
more can be imputed, based on these methods.

\begin{picture}(0,0)
\put(180,-50){\includegraphics[]{fig/impute_lr.pdf}}
\end{picture}






















