---
title: 'Exercises: error localization'
author: "Mark van der Loo"
date: "Statistics Netherlands"
output: 
  pdf_document:
    includes: 
      in_header: ../00tex/header.tex
params:
  answers: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Before we start.

1. In RStudio go to `tools->global options`. Under `Workspace`, make sure that
the option "`Restore .RData into workspace at startup`" has been switched off.
2. Press `<CTRL>+<SHIFT>+<F10>` to start a clean R session (in the same project).


We will keep using the `SBS2000.csv` dataset. So read it again.
```{r, echo=params$answers}
SBS2000 <- read.csv("SBS2000.csv", stringsAsFactors=FALSE)
```

\begin{picture}(0,0)
\put(350,-30){
\includegraphics[height=3cm]{fig/rstudio.png}
}
\end{picture}


#### Exercise 1

The `errorlocate` package has two principal functions:

- `locate_errors()` Localize errors based on the FH principle.
- `replace_errors()` Localize errors and replace them with other (or missing) values.

Load the `errorlocate` package.
```{r, messages=FALSE, results='hide'}
library(errorlocate)
```

a. Use `validator` to load the rules in `SBSrules.R` as `rules`
b. Print `rules` to see the validation rule set.
```{r, echo=params$answers, fig.show='hide', results='hide'}
rules <- validator(.file='SBSrules.R')
rules
```
c. Try the following code.
```{r}
el <- locate_errors(SBS2000, rules)
```
d. Use `summary` to summarize the results. Which variable is deemed erroneous
most often?


#### Exercise 2: weights.

Experts have indicated the following reliability weights for the SBS2000 variables.

\begin{tabular}{ll|ll}
staff       & 1   & turnover  & 1\\
other.rev   & 0.5 & total.rev & 1\\
staff.costs & 0.5 & profit    & 0.5\\
vat         & 1
\end{tabular}

a. Use these weights in error localization (see `?locate_errors`). Note: you
need to supply a weight vector of length `ncol(SBS2000)`. A simple option is to
add weights equal to 1 for the variables that are not checked. Store the result as
`el2`.
```{r, echo=params$answers, results='hide'}
weight <- rep(1,ncol(SBS2000))
names(weight) <- names(SBS2000)
weight[c("other.rev","staff.costs","profit")] <- 0.5

el2 <- locate_errors(SBS2000, rules, weight=weight)
summary(el2)
```
b. Compare the `summary` of `el` and `el2` are different variables
assigned as erroneous?
c. Use the following scaling rule to transform the weights, locate errors again, and compare your results.
$$
w'_j = 1 + \frac{w_j-w^{min}}{w^{max}-w^{min}}\times\frac{1}{n}
$$

```{r,echo=params$answers, results='hide'}
wmin <- min(weight)
wmax <- max(weight)
weight1 <- 1 + (weight-wmin)/(wmax-wmin)*1/length(weight)
# Note, we could reduce 'n' here, since the nr of variables
# occurring in the rules is smaller than the nr of variables 
# in the dataset. 
el3 <- locate_errors(SBS2000, rules, weight=weight1)
summary(el3)
```

### Exercise 3. Some rule set manipulations

It is possible to select rules by name or number, for example, this selects only the first
three rules and stures them in a new `validator` object, called `less_rules`. (try the code and check the result)

```{r,eval=TRUE, results='hide'}
less_rules <- rules[1:3]
```

It is also possible to add to `validator` objects together to create a new one.
Here is a simple example using two new validators. (try the code and check the result)
```{r, results='hide'}
v <- validator(x >= 0) + validator(x <= 100)
v
```

The function `variables` tells you what variables occurr in a rule set, for example. (try the code and check the result)
```{r, results='hide'}
variables(rules)
```


a. Which variables are occur in rules $1,3,5,7$?
b. List all variables in `SBS2000` that do _not occur_ in the rule set.
c. Amongst others, the variable `vat` does not occur in the rules. Add a rule
that states that VAT cannot be less than zero.
d. Export the new rule set to a `YAML` file, using `export_yaml` (see the help file).
e. Read the help file of `?variables` and answer the following question: which variable occurs in most rules?











