---
title: "Error localization"
author: "Mark van der Loo, Statistics Netherlands"
date: "Complutense University of Madrid, Spring 2019"
output: 
  beamer_presentation:
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---




## Error localization

Data validation and error localization answer different questions.

### Data validation

Which errors are there? 

### Error localization

Where do I need to make changes to fix the errors?


## Example

### Ruleset

```{r,eval=FALSE}
age >= 0
age <= 120
if (drivers_licence == TRUE) age >= 18
```

### Data

```{r, echo=FALSE}
knitr::kable(data.frame(age=10,drivers_licence=TRUE))
```

### Question:

Which field or fields would you change?



## Error localization

### Definition

Error localization is a procedure that points out fields in a
data set that can be altered or imputed in such a way that
all validation rules can be satisfied.


## Example

### Ruleset

```{r, eval=FALSE}
if (married == TRUE ) age >= 16
if (attends == "kindergarten") age <= 6
```

### Data

```{r, echo=FALSE}
knitr::kable(data.frame(age=3, married=TRUE, attends="kindergarten"))
```


### Question

Which field or fields would you change?

## Principle of Fellegi and Holt

###

Find the minimal (weighted) number of fields to adjust such that
all rules, including implied rules, can be satisfied.

\hfill{}\scriptsize{IP Fellegi and D Holt, JASA \textbf{71} 353  17--35 (1976).}

### Note

This should be used as a last resort, when no further information on 
the location of errors is available.


## Implied rules?

### 

```{r, eval=FALSE}
turnover - total.cost == profit
               profit <= 0.6 * turnover
```

This implies (substituting profit):

```{r, eval=FALSE}
           total.cost >= 0.4 * turnover
```

We need to take into account such _essentially new_ rules (edits) ---unstated 
relations between variables that can be derived from the explicitly defined rules.


## Error localization is a set covering problem

Given a record $\boldsymbol{x}=(x_1,\ldots,x_n)$, restricted by a set of
explicit and implicit validation rules. Minimize the sum

$$
\sum_{j\in S} w_j\textrm{, } w_j>0,
$$
Where $S\subseteq \{1,\ldots,n\}$ such that all (explicit and implicit)
validation rules can be satisfied by replacing $x_j, j\in S$ with new values.


\hfill{}\scriptsize{Algorithms: De Waal \emph{et al} (2011), John Wiley \& Sons.}\normalsize{}

## Solution types

### (1) Feasible solution

Any set of variables that can be altered so that all rules, including implied
ones can be satisfied. Example: change every variable.

### (2) Feasible solution of minimal size

A set of variables that is a feasible solution and as small as possible.

### (3) Feasible solution of minimal weight

A set of variables that is a feasible solution and minimizes the total weight.


## Choosing weights

### All weights equal (usually to one)

Least nr of variables adapted. In case of multiple solutions: choose randomly
(e.g. by adding a small random perturbation to the weights).

### Weights represent reliability

Heigher weight $\to$ variable is less likely chosen.

- Can be made to depend on 'outlierness', or expert judgement.
- Possible problem: minimal weights vs minimal nr of variables?


## Choosing weights

### Question

Is it possible to choose a set of weights, such that

a. The smallest number of variables is chosen
b. The weights are minimized


### Intuition

If the weights do not differ too much, no extra variables
will be introduced on top of the variables in a feasible
solution of of minimal size.


## Proposition 

Given a set of weights $\boldsymbol{w}=(w_1,\ldots,w_n)$ and write
$$
w_j = 1 + \delta_j\textrm{ with } 0\leq \delta_j\leq \delta^{max}.
$$
If $\delta^{max}<1/(n-1)$ then a feasible solution of minimal weight is
a feasible solution of minimal size.

\hfill{}\scriptsize{}Proof: van der Loo (2015) Rom. Stat. Rev. \textbf{2} 141--152; SDCR \S 7.5\normalsize{}


## Scaling rule

Given set of $n$ weights $\boldsymbol{w}=\boldsymbol{1}+\boldsymbol{\delta}$.
Then the following rescaling ensures that a solution of minimal weight also a
solution of minimal size.

$$
w'_j = 1 + \frac{w_j-w^{min}}{w^{max}-w^{min}}\times\frac{1}{n}
$$

### Procedure

1. Determine reliability weights (outlierness, expert judgement$\ldots$)
2. Apply scaling rule












